Fitness tracking reveals task-specific associations between memory, mental health, and physical activity

### 2022-07-13 paper/main.tex
working on addressing r1 comments

\title{Fitness tracking reveals task-specific associations between
  memory, mental health, and exercise}
\author[1, $\star$]{Jeremy R. Manning}
\author[1]{Gina M. Notaro}
\author[1]{Esme Chen}
\author[1]{Paxton C. Fitzpatrick}
\affil[1]{Dartmouth College, Hanover, NH}
\affil[$\star$]{Address correspondence to
  jeremy.r.manning@dartmouth.edu}

@@

\newpage
\begin{abstract}
  Physical exercise can benefit both physical and mental well-being.
  Different forms of exercise (e.g., aerobic versus anaerobic; running
  versus walking, swimming, or yoga; high-intensity interval training
  versus endurance workouts; etc.) impact physical fitness in
  different ways.  For example, running may substantially impact leg
  and heart strength but only moderately impact arm strength. We
  hypothesized that the mental benefits of exercise might be similarly
  differentiated.  We focused specifically on how different forms of
  exercise might relate to different aspects of memory and mental
  health.  To test our hypothesis, we collected (in aggregate) roughly
  a century's worth of fitness data.  We then asked participants to
  fill out surveys asking them to self-report on different aspects of
  their mental health.  We also asked participants to engage in a
  battery of memory tasks that tested their short and long term
  episodic, semantic, and spatial memory performance.  We found that
  participants with similar exercise habits and fitness profiles
  tended to also exhibit similar mental health and task performance
  profiles.  These effects were task-specific in that different
  exercise patterns or fitness characteristics varied with different
  aspects of memory, on different tasks.  Taken together, these
  findings provide foundational work for designing exercise
  interventions that target specific components of cognitive performance
  and mental health by leveraging low-cost fitness tracking devices.
\end{abstract}

\section*{Introduction}
Engaging in physical activity (exercise) can improve physical fitness
by increasing muscle strength \citep{RogeEvan93, Lind79, CranEtal13,
  Knut07}, bone density~\citep{ChilEtal12, BassRams94, LaynNels99},
cardiovascular performance~\citep{MaioEtal00, PollEtal00}, lung
capacity~\citep[][although see \cite{RomaEtal16}]{LazoEtal16}, and
endurance~\citep{WilmKnut03}.  Exercise can also improve mental
health~\citep{Ragl90, MikkEtal17, TaylEtal85, DeslEtal09, Call04,
  PaluSchw00, BassSuzu17} and cognitive
performance~\citep{ChanEtal12b, BrisEtal02, EtniEtal06, BassSuzu17}.

The physical benefits of exercise can be explained by stress-responses
of the affected body tissues. For example, skeletal muscles that are
taxed during exercise exhibit stress responses~\citep{MortEtal09} that
can in turn affect their growth or atrophy~\citep{SchiEtal13}.  By
comparison, the benefits of exercise on mental health are less direct.
For example, one hypothesis is that exercise leads to specific
physiological changes, such as increased aminergic synaptic
transmission and endorphin release, which in turn act on
neurotransmitters in the brain~\citep{PaluSchw00}.  Speculatively, if
different exercise regimens lead to different neurophysiological
responses, one might be able to map out a spectrum of signalling and
transduction pathways that are impacted by a given type, duration, and
intensity of exercise in each brain region.  For example, prior work
has shown that exercise increases acetylcholine levels, starting in
the vicinity of the exercised muscles~\citep{ShoeEtal97}.
Acetylcholine is thought to play an important role in memory
formation~\citep[e.g., by modulating specific synaptic inputs from
entorhinal cortex to the hippocampus, albeit in
rodents;][]{PalaEtal21}.  Given the central role that these medial
temporal lobe structures play in memory, changes in acetylcholine
might lead to specific changes in memory formation and retrieval.

In the present study, we hypothesize that (a) different exercise
regimens will have different, quantifiable impacts on cognitive
performance and mental health, and that (b) these impacts will be
consistant across individuals.  To this end, we collected a year of
real-world fitness tracking data from each of 113 participants.  We
then asked each participant to fill out a brief survey in which they
self-evaluated and self-reported several aspects of their mental
health.  Finally, we ran each participant through a battery of memory
tasks, which we used to evaluate their memory performance along
several dimensions.  We searched the data for potential associations
between memory, mental health, and exercise.

@@

    We ran an online experiment using the Amazon Mechanical Turk (MTurk)
    platform~\citep{GureEtal16}.  We collected data about each participant's fitness and
    exercise habits, a variety of self-reported measures concerning their
    mental health, and about their performance on a battery of memory
    tasks.

@@

\paragraph*{Survey questions.}  We collected the following demographic
information from each participant: their birth year, gender, highest
(academic) degree achieved, race, language fluency, and language
familiarity.  We also collected information about participants'
health and wellness, including about their vision, alertness, stress, sleep, coffee
and water consumption, location of their residence, activity typically
required for their job, and exercise habits.

@@

\section*{Results}
Before testing our main hypotheses, we examined the behavioral data
from each of four memory tasks: a random word list learning ``free
recall'' task (Fig.~\ref{fig:tasks}); a naturalistic recall task whereby participants watched
a short video and then recounted the narrative; a foreign language
``flashcards'' task; and a spatial learning task.  Each of the first
three tasks (free recall, naturalistic recall, and the flashcards
task) included both an immediate (short term) memory test and a
delayed (long term) memory test.  The spatial learning task included
only an immediate test. Participants in all four tasks exhibited
general trends and tendencies that have been previously reported in
prior work.  We were also interested in characterizing the variability
in task performance across participants.  For example, if all
participants exhibited near-identical behaviors or performance on a
given task, we would be unable to identify how memory performance on
that task varied with mental health or exercise.

@@

We wondered about potential links between the different aspects of
participants' data.  For example, if people who exercised in a
particular way also tended to perform better on a given memory task,
this could suggest that either (a) some property intrinsic to
participants who exercised in a particular way might also affect their
memory performance on the given task, and/or (b) particular exercise
behaviors could have a causal impact on memory performance.  We
carried out an exploratory analysis whereby we used a bootstrap-based
approach to identify reliable correlations between different aspects
of memory performance (Fig.~\behBehCorr), different aspects of fitness
(Fig.~\fitFitCorr), different demographic attributes
(Fig.~\demDemCorr), and correlations between memory performance,
fitness information, and demographic attributes (Fig.~\allCorr).
Specifically, we sought to identify correlations that were present in
the same direction (i.e., positive or negative) across different
subsets of participants.  For each test, we report the average correlation (taken
across 10,000 subsets of participants, chosen with replacement) and
an associated two-tailed $p$-value, estimated as
\[
p = 2 \times (1 - q),
\]
where $q$ is the proportion of those
10,000 subsets that exhibited correlations in the same direction  (see
\textit{Exploratory correlation analyses}).  When all 10,000 randomly
chosen subsets of participants exhibited correlations in the same
direction (i.e., all positive correlations or all negative
correlations), we report the $p$-value as $p < 0.0001$.

@@

The above analyses indicate that recent differences in fitness-related
activity are associated with differences in memory performance and
mental health measures.  Although the analyses treated these measures
on average or in aggregate, many of the measures we collected are
dynamic. For example, the amount or intensity of physical exercise
people engage in can vary over time, and so on.  We wondered whether
the dynamics of fitness-related measures might relate to memory
performance and/or mental health measures.  To this end, we carried out a
series of reverse correlation analyses (see \textit{Reverse
  correlation analyses}) to examine whether participants with
different cognitive or mental health profiles also tended to display
differences in fitness-related measures over time.  In particular, we
examined fitness data collected from participants' Fitbit devices over
the year prior to their test day in our study.  Several example findings are summarized in
Figure~\ref{fig:dynamics}.  We found that participants who performed
well on the immediate and delayed free recall memory tests and on the
naturalistic recall tests tended to be more active than participants
who performed poorly on those tests (Figs.~\ref{fig:dynamics}a, b;
\activityTimecourse).  Conversely, participants who performed well on
the immediate and delayed foreign language flashcards tasks tended to
be \textit{less} active.  These differences were present even a full
year before the testing day.  We also found substantial variability
across people with different (self-reported) mental health profiles
(Figs.~\ref{fig:dynamics}c, \activityTimecourseMH).  Due to small sample
sizes of individuals exhibiting several mental health dimensions, it
is difficult to distinguish generalizable trends from individual
differences that one or two individuals happened to exhibit.  However,
several large-sample-size trends emerged.  For example, participants
who reported higher levels of stress also tended to be slightly more
physically active than participants who reported lower stress levels.
We found analogous differences in other activity-related measures
(Figs.~\activityTimecourse~and \activityTimecourseMH), cardiovascular
measures (Figs.~\cardioTimecourse~and \cardioTimecourseMH), and
sleep-related measures (Figs.~\sleepTimecourse~and
\sleepTimecourseMH).  Taken together, the analyses suggest that
cognitive and mental health differences are also associated with differences
in the dynamics of physical health measures.

@@

One important limitation of our study is that we cannot distinguish
correlations between different measures from potential causal
effects.  For example, we cannot know (from our study) whether
engaging in particular forms of exercise \textit{causes} changes in
memory performance or mental health, or whether (alternatively) people
who tend to engage in similar forms of exercise also happen to
exhibit similar memory and/or mental health profiles.  In other words,
an overlapping set of processes or person-specific attributes may lead
someone to both form particular habits around exercise and display high or low performance on a
given memory test.  We do not know whether memory performance or
aspects of mental health might be manipulated or influenced by
changing the patterns of physical activity someone engages in.  For
this reason, we have been careful to frame our findings as
correlations and associations, rather than to imply knowledge about a causal direction
of our findings.

Although the present study cannot reveal causal effects, a large prior
literature provides some insight into potential causal effects by examining the neural and cognitive effects of a variety
of exercise interventions~\citep{ChanEtal15, VidoEtal15, KamiEtal07,
  ImboEtal19, SuwaEtal17, SinhEtal21, Tomp03}.  A limitation of that
prior work is that most of these studies examine how relatively
short-term changes in exercise (e.g., on timescales of hours to days
or, rarely, weeks to months) affect a cognitive performance on single
task or aspect of mental health.  The present study examines
longer-term exercise (over a full year), and relates long-term
exercise history to performance on a variety of tasks and to a variety
of mental health dimensions.

To the extent that exercise \textit{does} provide a non-invasive means
of manipulating cognitive performance and mental health, our work may
have exciting implications for cognitive enhancement.  For example,
one might imagine building a recommendation system that suggests a
particular exercise regimen tailored to improve a specific aspect of
an individual's cognitive performance (e.g., the efficacy of a student's study session for an
upcoming exam) or mental health (e.g., reducing symptoms of anxiety
before an important meeting).  Just as strength training may be
customized to target a specific muscle group, or to improve
performance on a specific physical task, similar principles might also
be applied to target specific improvements in cognitive fitness
and mental health.

@@

\bibliographystyle{apa}
\bibliography{/Users/jmanning/CDL-bibliography/cdl}
\end{document}

>>>>

\title{Fitness tracking reveals task-specific associations between
  memory, mental health, and physical activity}
\author[1, $\star$]{Jeremy R. Manning}
\author[1]{Gina M. Notaro}
\author[1]{Esme Chen}
\author[1]{Paxton C. Fitzpatrick}
\affil[1]{Dartmouth College, Hanover, NH}
\affil[$\star$]{Address correspondence to
  jeremy.r.manning@dartmouth.edu}

@@

\newpage
\begin{abstract}
  Physical activity can benefit both physical and mental well-being.
  Different forms of exercise (e.g., aerobic versus anaerobic; running
  versus walking, swimming, or yoga; high-intensity interval training
  versus endurance workouts; etc.) impact physical fitness in
  different ways.  For example, running may substantially impact leg
  and heart strength but only moderately impact arm strength. We
  hypothesized that the mental benefits of physical activity might be similarly
  differentiated.  We focused specifically on how different intensities of
  physical activity might relate to different aspects of memory and mental
  health.  To test our hypothesis, we collected (in aggregate) roughly
  a century's worth of fitness data.  We then asked participants to
  fill out surveys asking them to self-report on different aspects of
  their mental health.  We also asked participants to engage in a
  battery of memory tasks that tested their short and long term
  episodic, semantic, and spatial memory performance.  We found that
  participants with similar physical activity habits and fitness profiles
  tended to also exhibit similar mental health and task performance
  profiles.  These effects were task-specific in that different
  physical activity patterns or fitness characteristics varied with different
  aspects of memory, on different tasks.  Taken together, these
  findings provide foundational work for designing physical activity
  interventions that target specific components of cognitive performance
  and mental health by leveraging low-cost fitness tracking devices.
\end{abstract}

\section*{Introduction}
Engaging in physical activity (exercise) can improve physical fitness
by increasing muscle strength \citep{RogeEvan93, Lind79, CranEtal13,
  Knut07}, bone density~\citep{ChilEtal12, BassRams94, LaynNels99},
cardiovascular performance~\citep{MaioEtal00, PollEtal00}, lung
capacity~\citep[][although see \cite{RomaEtal16}]{LazoEtal16}, and
endurance~\citep{WilmKnut03}.  Physical activity can also improve mental
health~\citep{Ragl90, MikkEtal17, TaylEtal85, DeslEtal09, Call04,
  PaluSchw00, BassSuzu17} and cognitive
performance~\citep{ChanEtal12b, BrisEtal02, EtniEtal06, BassSuzu17}.

The physical benefits of exercise can be explained by stress-responses
of the affected body tissues. For example, skeletal muscles that are
taxed during exercise exhibit stress responses~\citep{MortEtal09} that
can in turn affect their growth or atrophy~\citep{SchiEtal13}.  By
comparison, the benefits of physical activity on mental health are less direct.
For example, one hypothesis is that physical activity leads to specific
physiological changes, such as increased aminergic synaptic
transmission and endorphin release, which in turn act on
neurotransmitters in the brain~\citep{PaluSchw00}.  Speculatively, if
different physical activity regimens lead to different neurophysiological
responses, one might be able to map out a spectrum of signalling and
transduction pathways that are impacted by a given type, duration, and
intensity of physical activity in each brain region.  For example, prior work
has shown that physical activity increases acetylcholine levels, starting in
the vicinity of the exercised muscles~\citep{ShoeEtal97}.
Acetylcholine is thought to play an important role in memory
formation~\citep[e.g., by modulating specific synaptic inputs from
entorhinal cortex to the hippocampus, albeit in
rodents;][]{PalaEtal21}.  Given the central role that these medial
temporal lobe structures play in memory, changes in acetylcholine
might lead to specific changes in memory formation and retrieval.

In the present study, we hypothesize that (a) different exercise
regimens will have different, quantifiable impacts on cognitive
performance and mental health, and that (b) these impacts will be
consistant across individuals.  To this end, we collected a year of
real-world fitness tracking data from each of 113 participants.  We
then asked each participant to fill out a brief survey in which they
self-evaluated and self-reported several aspects of their mental
health.  Finally, we ran each participant through a battery of memory
tasks, which we used to evaluate their memory performance along
several dimensions.  We searched the data for potential associations
between memory, mental health, and physical activity.

@@

    We ran an online experiment using the Amazon Mechanical Turk (MTurk)
    platform~\citep{GureEtal16}.  We collected data about each participant's fitness and
    physical activity habits, a variety of self-reported measures concerning their
    mental health, and about their performance on a battery of memory
    tasks.

@@

\paragraph*{Survey questions.}  We collected the following demographic
information from each participant: their birth year, gender, highest
(academic) degree achieved, race, language fluency, and language
familiarity.  We also collected information about participants'
health and wellness, including about their vision, alertness, stress, sleep, coffee
and water consumption, location of their residence, activity typically
required for their job, and physical activity habits.

@@

\section*{Results}
Before testing our main hypotheses, we examined the behavioral data
from each of four memory tasks: a random word list learning ``free
recall'' task (Fig.~\ref{fig:tasks}); a naturalistic recall task whereby participants watched
a short video and then recounted the narrative; a foreign language
``flashcards'' task; and a spatial learning task.  Each of the first
three tasks (free recall, naturalistic recall, and the flashcards
task) included both an immediate (short term) memory test and a
delayed (long term) memory test.  The spatial learning task included
only an immediate test. Participants in all four tasks exhibited
general trends and tendencies that have been previously reported in
prior work.  We were also interested in characterizing the variability
in task performance across participants.  For example, if all
participants exhibited near-identical behaviors or performance on a
given task, we would be unable to identify how memory performance on
that task varied with mental health or physical activity.

@@

We wondered about potential links between the different aspects of
participants' data.  For example, if people who exercised in a
particular way also tended to perform better on a given memory task,
this could suggest that either (a) some property intrinsic to
participants who exercised in a particular way might also affect their
memory performance on the given task, and/or (b) particular physical activity
behaviors could have a causal impact on memory performance.  We
carried out an exploratory analysis whereby we used a bootstrap-based
approach to identify reliable correlations between different aspects
of memory performance (Fig.~\behBehCorr), different aspects of fitness
(Fig.~\fitFitCorr), different demographic attributes
(Fig.~\demDemCorr), and correlations between memory performance,
fitness information, and demographic attributes (Fig.~\allCorr).
Specifically, we sought to identify correlations that were present in
the same direction (i.e., positive or negative) across different
subsets of participants.  For each test, we report the average correlation (taken
across 10,000 subsets of participants, chosen with replacement) and
an associated two-tailed $p$-value, estimated as
\[
p = 2 \times (1 - q),
\]
where $q$ is the proportion of those
10,000 subsets that exhibited correlations in the same direction  (see
\textit{Exploratory correlation analyses}).  When all 10,000 randomly
chosen subsets of participants exhibited correlations in the same
direction (i.e., all positive correlations or all negative
correlations), we report the $p$-value as $p < 0.0001$.

@@

The above analyses indicate that recent differences in fitness-related
activity are associated with differences in memory performance and
mental health measures.  Although the analyses treated these measures
on average or in aggregate, many of the measures we collected are
dynamic. For example, the amount or intensity of physical activity
people engage in can vary over time, and so on.  We wondered whether
the dynamics of fitness-related measures might relate to memory
performance and/or mental health measures.  To this end, we carried out a
series of reverse correlation analyses (see \textit{Reverse
  correlation analyses}) to examine whether participants with
different cognitive or mental health profiles also tended to display
differences in fitness-related measures over time.  In particular, we
examined fitness data collected from participants' Fitbit devices over
the year prior to their test day in our study.  Several example findings are summarized in
Figure~\ref{fig:dynamics}.  We found that participants who performed
well on the immediate and delayed free recall memory tests and on the
naturalistic recall tests tended to be more active than participants
who performed poorly on those tests (Figs.~\ref{fig:dynamics}a, b;
\activityTimecourse).  Conversely, participants who performed well on
the immediate and delayed foreign language flashcards tasks tended to
be \textit{less} active.  These differences were present even a full
year before the testing day.  We also found substantial variability
across people with different (self-reported) mental health profiles
(Figs.~\ref{fig:dynamics}c, \activityTimecourseMH).  Due to small sample
sizes of individuals exhibiting several mental health dimensions, it
is difficult to distinguish generalizable trends from individual
differences that one or two individuals happened to exhibit.  However,
several large-sample-size trends emerged.  For example, participants
who reported higher levels of stress also tended to be slightly more
physically active than participants who reported lower stress levels.
We found analogous differences in other activity-related measures
(Figs.~\activityTimecourse~and \activityTimecourseMH), cardiovascular
measures (Figs.~\cardioTimecourse~and \cardioTimecourseMH), and
sleep-related measures (Figs.~\sleepTimecourse~and
\sleepTimecourseMH).  Taken together, the analyses suggest that
cognitive and mental health differences are also associated with differences
in the dynamics of physical health measures.

@@

One important limitation of our study is that we cannot distinguish
correlations between different measures from potential causal
effects.  For example, we cannot know (from our study) whether
engaging in particular forms of physical activity \textit{causes} changes in
memory performance or mental health, or whether (alternatively) people
who tend to engage in similar forms of physical activity also happen to
exhibit similar memory and/or mental health profiles.  In other words,
an overlapping set of processes or person-specific attributes may lead
someone to both form particular habits around physical activity and display high or low performance on a
given memory test.  We do not know whether memory performance or
aspects of mental health might be manipulated or influenced by
changing the patterns of physical activity someone engages in.  For
this reason, we have been careful to frame our findings as
correlations and associations, rather than to imply knowledge about a causal direction
of our findings.

Although the present study cannot reveal causal effects, a large prior
literature provides some insight into potential causal effects by examining the neural and cognitive effects of a variety
of exercise interventions~\citep{ChanEtal15, VidoEtal15, KamiEtal07,
  ImboEtal19, SuwaEtal17, SinhEtal21, Tomp03}.  A limitation of that
prior work is that most of these studies examine how relatively
short-term changes in physical activity (e.g., on timescales of hours to days
or, rarely, weeks to months) affect a cognitive performance on single
task or aspect of mental health.  The present study examines
longer-term physical activity (over a full year), and relates long-term
physical activity history to performance on a variety of tasks and to a variety
of mental health dimensions.

To the extent that physical activity \textit{does} provide a non-invasive means
of manipulating cognitive performance and mental health, our work may
have exciting implications for cognitive enhancement.  For example,
one might imagine building a recommendation system that suggests a
particular physical activity regimen tailored to improve a specific aspect of
an individual's cognitive performance (e.g., the efficacy of a student's study session for an
upcoming exam) or mental health (e.g., reducing symptoms of anxiety
before an important meeting).  Just as strength training may be
customized to target a specific muscle group, or to improve
performance on a specific physical task, similar principles might also
be applied to target specific improvements in cognitive fitness
and mental health.

@@

\bibliographystyle{apa}
\bibliography{/home/jeremy/CDL-bibliography/cdl.bib}
\end{document}
